---
title: "Weather Event and the impact on population"
author: "Abhishek Sinha"
date: "July 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2, quietly = TRUE)
```

# Impact of weather events on population health and economy

## Synopsis

With the data provided by U.S. National Oceanic and Atmospheric Administration on storm, we have performed analysis on the impact of weather event on population health(fatalities and injuries) and economy.  

The results of the analysis points to Tornado and Hail as the biggest weather events to have an impact on health and economy. Tornado has had the highest fatalities and injuries.

Similarly, Hail has impacted the most on Crop Damages, which Tornado has biggest effect on Property Damages.

## Loading and processing the data

```{r cache=TRUE}
if(!file.exists("StormData.csv.bz2"))
{
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","StormData.csv.bz2")
}
storm.data <- read.csv("StormData.csv.bz2", header = T)
```


```{r}
storm.data <- storm.data[,c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
```

### Function to replace units

Units (H, K, M and B) are defined in the two fields (PROPDMGEXP, CROPDMGEXP). It is important to convert them into numbers for comparision. E.g. H - 100, K - 1000, M - 1000000, B - 1000000000

```{r}
replace_units <- function(x)
{
        ifelse(is.na(x),return(1),
               ifelse(x=='K'|x=='k',return(1000),
                      ifelse(x=='H'|x=='h',return(100),
                             ifelse(x=='M'|x=='m',return(1000000),
                                    ifelse(x=='B'|x=='b',return(1000000000),1)
                             )
                      )
               )
        )
        return(1)
}
```

### Tidying up data

Update Crop Damage and Property damage appropriately.

```{r}
storm.data$PROPDMGEXP <- replace_units(storm.data$PROPDMGEXP)
storm.data$CROPDMGEXP <- replace_units(storm.data$CROPDMGEXP)
storm.data$PROPDMG <- storm.data$PROPDMG * storm.data$PROPDMGEXP
storm.data$CROPDMG <- storm.data$CROPDMG * storm.data$CROPDMGEXP
storm.data$PROPDMGEXP <- NULL
storm.data$CROPDMGEXP <- NULL

#Update column names of the dataframe
colnames(storm.data) <- c("EventType","Fatalities","Injuries","PropertyDamage","CropDamage")
```

--------------

## Events that are most harmful to population health

### Fatalities  
Summarize the sum of fatalities by event type and look at top 10 events

```{r}
fatalitiest.eventtype <- aggregate(Fatalities ~ EventType, data = storm.data, FUN = sum)
top.event.fatalities <- head(fatalitiest.eventtype[order(-fatalitiest.eventtype$Fatalities),],10)
```


```{r}
ggplot(top.event.fatalities) + aes(x = EventType, y=Fatalities, fill=EventType) + geom_bar(stat="identity") + 
    xlab("Weather Events") + ylab("Fatalities") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### Injuries  
Summarize the sum of injuries by event type and look at top 10 events

```{r}
injuries.eventtype <- aggregate(Injuries ~ EventType, data = storm.data, FUN = sum)
top.event.injuries <- head(injuries.eventtype[order(-injuries.eventtype$Injuries),],10)
```


```{r}
ggplot(top.event.injuries) + aes(x = EventType, y=Injuries, fill=EventType) + geom_bar(stat="identity") + 
    xlab("Weather Events") + ylab("Injuries") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


--------------

## Events that had the greatest economic consequences

### Property Damage  
Summarize the sum of property damage by event type and look at top 10 events

```{r}
property.damage.eventtype <- aggregate(PropertyDamage ~ EventType, data = storm.data, FUN = sum)
top.event.prop.damage <- head(property.damage.eventtype[order(-property.damage.eventtype$PropertyDamage),],10)
head(top.event.prop.damage, 10)
```

### Crop Damage  
Summarize the sum of crop damage by event type and look at top 10 events

```{r}
crop.damage.eventtype <- aggregate(CropDamage ~ EventType, data = storm.data, FUN = sum)
top.event.crop.damage <- head(crop.damage.eventtype[order(-crop.damage.eventtype$CropDamage),],10)
head(top.event.crop.damage, 10)
```

### Overall Damage    
Summarize the sum of amount of crop and food damage by event type and look at top 5 events

```{r}
storm.data$CropPropertyDamage <- storm.data$CropDamage + storm.data$PropertyDamage
damage.eventtype <- aggregate(CropPropertyDamage ~ EventType, data = storm.data, FUN = sum)
top.event.damage <- head(damage.eventtype[order(-damage.eventtype$CropPropertyDamage),],10)
```


```{r}
ggplot(top.event.damage) + aes(x = EventType, y=CropPropertyDamage, fill=EventType) + geom_bar(stat="identity") + 
    xlab("Weather Events") + ylab("Crop and Property Damage") + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


## Results

The final analysis suggests Tornado being the biggest impactful weather event impacting both health and economy. TSTM Wind, Flash Flood and Hail also has a high impact on economy.

Similarly, Tornado is the highes fatalities and has a similar impact on number of injuries.